Solution of an Acoustic Transmission Inverse Problem by Extended Inversion

### 2020-11-23 repro/papers/discrep/paper.tex
interim

In the next (``Overview'') section, we describe the conceptual
framework of this inverse problem, review the related literature, and
state the essence of our main results, leaving out a few technical
details that are supplied later. Numerical illustrations of these
results occupy the following (``Examples'') section. The fourth
(``Findings'') section of the paper contains precise statements of the
mathematical results and supporting discussion, followed by
derivations. In particular, theorems referenced in the Overview
section are to be found, stated in full with their proofs, in the this
final part of the paper. Three appendices describe the algorithm used to
control the penalty weight in our formulation of extended
inversion, further refinement of error estimates based ond the
Heisenberg inequality, and the abstract structure of objective derivatives
inherent in extended inversion based on wave propagation.

@@

\subsection{Results}
While $F[m]$ is surjective for every $m \in M$, as noted above, it is
very far from injective. On the other hand, under a constraint that
will be assumed throughout, $\lF[m]$ is injective (in fact, $4 \pi r
\lF[m]$ is an isometry):
\begin{proposition}
  \label{thm:fullrec}
  Suppose that 
  \begin{equation}
    \label{eqn:fullrec}
    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
    \subset [t_{\rm min},t_{\rm max}].
  \end{equation}
  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
  (0,\lambda_{\rm max}]$.
\end{proposition}

@@

The main theoretical device used in the proofs of our main results on 
extended inversion is Variable Projection reduction of the penalty objective $\Ja$
(equation \ref{eqn:penerr}) to a function $\tJa$ of 
$m$ alone, by minimization over $w$, as described in the Examples
section. We repeat the definition here for the reader's convenience:
\begin{equation}
  \label{eqn:redexp}
  \tJa[m;d] = \inf_w \Ja[m,w;d]
\end{equation}

@@

The choices for the penalty operator $A$ considered here are scalar 
multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
\begin{equation}
  \label{eqn:annmult}
  A w(t)= a(t)w(t), \, t\in \bR.
\end{equation}

@@

\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
were differentiable, then ``local minimizer'' in the conclusion of the
preceding theorem could be replaced by ``stationary point''. However,
for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
  not} differentiable without added smoothness constraints on $w$,
whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
penalty operator $A$:

@@

This phenomenon - nondifferentiable objective, differentiable reduced
objective - is replicated for a wide variety of extended
approaches for inverse problems in wave propagation. See for example \cite{Symes:EAGE15}.

@@

\subsection{Proof of Theorem \ref{thm:fwi}}
\noindent{\bf Remark:} A useful consequence of the condition 
\ref{eqn:fullrec}: for every $m \in M$, 
\begin{equation}
  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
  t_{\rm max}-mr]. 
  \label{eqn:zeroinc}
\end{equation}

@@

\begin{proof} of Theorem \ref{thm:fwi}: From the definition \ref{eqn:mod},
\[
 \lerr[m,w;d] =  \frac{1}{32\pi^2
    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
\]
Since $w_*, w$ vanish for $|t|>\lambda$,
$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
$\langle \cdot,\cdot \rangle_D$ on $D$:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
  F[m_*]w_*\rangle_D = 0
\end{equation}
But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
orthogonal to $F[m]w$. So conclude after a minor manipulation that
\[
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
\]
\begin{equation}
  \label{eqn:iso}
  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
\end{equation}
That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
$m$, and its minimum over $w$ is attained for $w=0$
\end{proof}
%\begin{theorem}
%  \label{thm:fwi}
%  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
%  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
%  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
%\begin{equation}
%  \label{eqn:isovpm}
% \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
%\end{equation}
%and any such $(m,0)$ is a stationary point of $e$.
%\end{theorem}

Therefore local minimizers of $\lerr$ abound, as far as you like from the
global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
$e$ gives no useful information whatever about constructive search
directions, and descent-based optimization tends to fail if the
initial estimate $m_0$ is in error by more than $2\lambda/r$
(``further than a multiple of a wavelength'', per discussion
above). In fact the actual behaviour of FWI itererations is worse
(failure if $m_0$ is in error by ``half a wavelength''), as follows
from a more refined analysis of the ``cycle-skipping'' local behaviour of $\lerr$ near its
global minimizer.

@@

\subsection{Proof of Theorem \ref{thm:norminv}}
For the single-trace transmission inverse problem, $\tilde{J}_{\alpha}$ is explicitly
computable. First observe that apart from amplitude, $F[m]$ is
unitary: for $g \in D$,
\begin{equation}
\label{eqn:tran}
F[m]^T g (t) =
\left\{
  \begin{array}{c}
    \frac{1}{4\pi r}g\left(t+mr\right), \, t \in [t_{\rm min}-mr,
    t_{\rm max}-mr],\\
    0, \mbox{ else.}
  \end{array}
\right.
\end{equation}
so
\begin{equation}
  \label{eqn:unit}
  F[m]^TF[m] = \frac{1}{(4\pi r)^2}{\bf 1}_{[t_{\rm min}-mr,  
    t_{\rm max}-mr]}
\end{equation}
in which ${\bf 1}_{S}$ denotes
multiplication by the characteristic function of a measurable 
$S \subset \bR$.

\begin{proof} of Theorem \ref{thm:norminv}:
  \begin{itemize}
  \item[1. ]Note that thanks to \ref{eqn:zeroinc}, if $|t|\le
    \lambda \le \lambda_{\rm max}$, then ${\bf 1}_{[t_{\rm min}-mr,  
      t_{\rm max}-mr]}(t) = 1$, whereas if $|t|>\lambda$,
    then $a(t) \ge C$, whence
    \[
      \frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
        t_{\rm max}-mr]} + \alpha^2a^2  \ge \min\{(4\pi r)^2,
      \alpha^2\min\{1/(4\pi r)^2,C^2\} \}> 0.
    \]
    Therefore the normal operator is invertible under the stated
    conditions.

@@

\begin{proof} of Theorem \ref{thm:diffobj}:
Suppose first that $a \in C^1(\bR)$. Differentiation under the integral sign  
  yields the expression for its derivative:
\begin{equation}
  \label{eqn:dexpjgen}
  \frac{d}{dm}\tJa[m;d] = -\frac{(4 \pi r \alpha)^2}{\|d\|^2} \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
  a(t-mr))^2)^{-2}d(t)^2. 
\end{equation}
For $a \in W^{1,\infty}_{\rm loc}(\bR)$ a limiting argument shows that the
same expression gives the derivative of $\tJa$.
\end{proof}

\subsection{Proof of Theorem \ref{thm:boxcarbad}}
\begin{corollary}
  \label{thm:epjnonoise}
  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
\begin{equation}
  \label{eqn:residnorm}
  e[m,\aw[m,d];d] 
= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
\end{equation}
\begin{equation}
  \label{eqn:anninorm}
  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
\end{equation}
so 
\begin{equation}
\label{eqn:expjnonoise}
\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
\end{equation}
Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
\begin{equation}
  \label{eqn:dexpjnonoise}
  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
\end{equation}
\end{corollary}

>>>>

In the next (``Overview'') section, we describe the conceptual
framework of this inverse problem, review the related literature, and
state the essence of our main results, leaving out a few technical
details that are supplied later. Numerical illustrations of these
results occupy the following (``Examples'') section. The fourth
(``Findings'') section of the paper contains precise statements of the
mathematical results, derivations, and supporting discussion. Three
appendices describe the algorithm used to control the penalty weight
in our formulation of extended inversion, further refinement of error
estimates based ond the Heisenberg inequality, and the abstract
structure of extended inversion implicit in the derivations presented
in the ``Findings'' section.

@@

\subsection{Full Waveform Inversion}
While $F[m]$ is surjective for every $m \in M$, as noted above, it is
very far from injective. On the other hand, under a constraint that
will be assumed throughout, $\lF[m]$ is injective (in fact, $4 \pi r
\lF[m]$ is an isometry):
\begin{proposition}
  \label{thm:fullrec}
  Suppose that 
  \begin{equation}
    \label{eqn:fullrec}
    [ m_{\rm min}r-\lambda_{\rm max}, m_{\rm max}r+\lambda_{\rm max}]
    \subset [t_{\rm min},t_{\rm max}].
  \end{equation}
  Then $\lF[m]\in {\cal B}(\lW,D)$ is coercive for every $m \in M, \lambda \in
  (0,\lambda_{\rm max}]$.
\end{proposition}

\noindent{\bf Remark:} A useful consequence of the condition 
\ref{eqn:fullrec}: for every $m \in M$, 
\begin{equation}
  [-\lambda_{\rm max}, -\lambda_{\rm max}] \subset[ t_{\rm min}-mr , 
  t_{\rm max}-mr]. 
  \label{eqn:zeroinc}
\end{equation}

@@

\begin{proof} From the definition \ref{eqn:mod},
\[
 \lerr[m,w;d] =  \frac{1}{32\pi^2
    r^2\|d\|^2}\int_0^T\,dt\,\left|w\left(t-mr\right)-w_*\left(t-m_*r\right)\right|^2
\]
Since $w_*, w$ vanish for $|t|>\lambda$,
$\lF[m_*]w_*(t)$ vanishes if $|t-m_*r|>\lambda$ and $\lF[m]w$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $\lF[m]w(t)=0$ or $\lF[m_*]w_*=0$. Therefore $\lF[m]w$ and
$\lF[m_*]w_*$ are orthogonal in the sense of the $L^2$ inner product
$\langle \cdot,\cdot \rangle_D$ on $D$:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle F[m]w,
  F[m_*]w_*\rangle_D = 0
\end{equation}
But $d = \lF[m_*]w_*$, so this is the same as saying that $d$ is
orthogonal to $F[m]w$. So conclude after a minor manipulation that
\[
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \lerr[m,w;d]=\frac{1}{32\pi^2 
    r^2\|d\|^2}(\|w\|^2 + \|w_*\|^2).
\]
\begin{equation}
  \label{eqn:iso}
  = \frac{1}{2}\left(\frac{\|w\|^2}{\|w_*\|^2} + 1 \right)
\end{equation}
That is, for slowness $m$ in error by more than $2\lambda/r$ from the 
target slowness $m_*$, the means square error (FWI objective) $\lerr$ is independent of
$m$, and its minimum over $w$ is attained for $w=0$
\end{proof}

Therefore local minimizers of $\lerr$ abound, as far as you like from the
global minimizer $(m_*,w_*)$. Local exploration of the FWI objective
$e$ gives no useful information whatever about constructive search
directions, and descent-based optimization tends to fail if the
initial estimate $m_0$ is in error by more than $2\lambda/r$
(``further than a multiple of a wavelength'', per discussion
in the second Appendix). In fact the actual behaviour of FWI itererations is worse
(failure if $m_0$ is in error by ``half a wavelength''), as follows
from a more refined analysis of the ``cycle-skipping'' local behaviour of $\lerr$ near its
global minimizer.

\subsection{Variable Projection}
The main theoretical device used in the proofs of our main results on 
extended inversion is Variable Projection reduction of the penalty objective $\Ja$
(equation \ref{eqn:penerr}) to a function $\tJa$ of 
$m$ alone, by minimization over $w$, as described in the Examples
section. We repeat the definition here for the reader's convenience:
\begin{equation}
  \label{eqn:redexp}
  \tJa[m;d] = \inf_w \Ja[m,w;d]
\end{equation}

@@

The choices for the penalty operator $A$ considered here are scalar 
multiplication operators on $W$ defined by a choice of multiplier $a \in L^{\infty}(\bR)$:
\begin{equation}
  \label{eqn:annmult}
  A w(t)= a(t)w(t), \, t\in \bR.
\end{equation}
With $A$ of this form, $\tilde{J}_{\alpha}$ is explicitly
computable. First observe that apart from amplitude, $F[m]$ is
unitary: for $g \in D$,
\begin{equation}
\label{eqn:tran}
F[m]^T g (t) =
\left\{
  \begin{array}{c}
    \frac{1}{4\pi r}g\left(t+mr\right), \, t \in [t_{\rm min}-mr,
    t_{\rm max}-mr],\\
    0, \mbox{ else.}
  \end{array}
\right.
\end{equation}
so
\begin{equation}
  \label{eqn:unit}
  F[m]^TF[m] = \frac{1}{(4\pi r)^2}{\bf 1}_{[t_{\rm min}-mr,  
    t_{\rm max}-mr]}
\end{equation}
in which ${\bf 1}_{S}$ denotes
multiplication by the characteristic function of a measurable 
$S \subset \bR$.

\begin{proof} 
  \begin{itemize}
  \item[1. ]Note that thanks to \ref{eqn:zeroinc}, if $|t|\le
    \lambda \le \lambda_{\rm max}$, then ${\bf 1}_{[t_{\rm min}-mr,  
      t_{\rm max}-mr]}(t) = 1$, whereas if $|t|>\lambda$,
    then $a(t) \ge C$, whence
    \[
      \frac{1}{(4\pi r)^2}  {\bf 1}_{[t_{\rm min}-mr,  
        t_{\rm max}-mr]} + \alpha^2a^2  \ge \min\{(4\pi r)^2,
      \alpha^2\min\{1/(4\pi r)^2,C^2\} \}> 0.
    \]
    Therefore the normal operator is invertible under the stated
    conditions.

@@

\begin{proof} These conclusions follow immediately from Proposition
  \ref{thm:norminvexp}.
\end{proof}

\noindent{\bf Remark:} If $\Ja[\cdot,\cdot;d]$ and $\tJa[\cdot;d]$
were differentiable, then ``local minimizer'' in the conclusion of the
preceding theorem could be replaced by ``stationary point''. However,
for the problem addressed in this paper, $\Ja[\cdot,\cdot;d]$ {\em is
  not} differentiable without added smoothness constraints on $w$,
whereas $\tJa[\cdot;d]$ {\em is} differentiable for proper choice of
penalty operator $A$.

This conclusion follows from properties of the modeling operator $F$
shared with many other inverse problems in wave propagation, as
explained in the third appendix. Here, we derive it from explicit
expressions for $\tJa$ and its components.

@@

\begin{proof}
Suppose first that $a \in C^1(\bR)$. Differentiation under the integral sign  
  yields the expression for its derivative:
\begin{equation}
  \label{eqn:dexpjgen}
  \frac{d}{dm}\tJa[m;d] = -\frac{(4 \pi r \alpha)^2}{\|d\|^2} \int_{t_{\rm min}}^{t_{\rm max}} \,dt \, 
  \left(a\frac{da}{dt}\right)(t-mr)(1+(4\pi r \alpha 
  a(t-mr))^2)^{-2}d(t)^2. 
\end{equation}
For $a \in W^{1,\infty}_{\rm loc}(\bR)$ a limiting argument shows that the
same expression gives the derivative of $\tJa$.
\end{proof}

It will be useful to record expressions for the various componenets of
$\tJa$ when the data is noise-free, that is, the context of
Proposition \ref{thm:norminvexp}, item 3.

\begin{corollary}
  \label{thm:epjnonoise}
  Assume the hypotheses of Proposition \ref{thm:norminvexp}, item
  3. Then noting that $\|d\| = \|w_*\|/(4 \pi r)$
\begin{equation}
  \label{eqn:residnorm}
  e[m,\aw[m,d];d] 
= \frac{\alpha^4}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^4(1+(4\pi r)^2 \alpha^2 
    a(t-(m-m^*)r)^2)^{-2}w_*(t)^2.
\end{equation}
\begin{equation}
  \label{eqn:anninorm}
  p[m,\aw[m,d];d] = \frac{(4\pi r)^2}{2\|w_*\|^2}\int \,dt\,  a(t-(m-m_*)r)^2 \left(1+ (4\pi r)^2\alpha^2
    a(t-(m-m_*)r)^2\right)^{-2}w_*(t)^2.
\end{equation}
so 
\begin{equation}
\label{eqn:expjnonoise}
\tJa[m;d] = \frac{(4\pi r \alpha)^2}{2\|w_*\|^2}\int\,dt\,a(t-(m-m_*)r)^2(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-1}w_*(t)^2. 
\end{equation}
Finally, if $a \in W^{1,1}_{\rm loc}(\bR)$, then $\tJa[\cdot;d]$ is differentiable, and 
\begin{equation}
  \label{eqn:dexpjnonoise}
  \frac{d}{dm}\tJa[m;d] = -\frac{r (4\pi r \alpha)^2}{\|w_*\|^2} \int \,dt \, 
  \left(a\frac{da}{dt}\right)(t-(m-m_*)r)(1+(4\pi r)^2 \alpha^2 
  a(t-(m-m_*)r)^2)^{-2}w_*(t)^2. 
\end{equation}
\end{corollary}

\subsection{Choice of Penalty Operator}

@@

\subsection{Proof of Theorem \ref{thm:fwi}}

@@

%\begin{theorem}
%  \label{thm:fwi}
%  Suppose that $0 <\lambda\le \lambda_{\rm max}$,  $m_* \in M, w_*
%  \in \lW, d=\lF[m_*]w_*$ is noise-free data per definition \ref{eqn:defdatanonoise},
%  Under assumption \ref{eqn:fullrec}, for any $m \in M$ with $r|m-m_*|>2\lambda$,
%\begin{equation}
%  \label{eqn:isovpm}
% \min_w e[m,w;d]=e[m,0,d]=\frac{1}{32\pi^2 r^2}\|w_*\|^2,
%\end{equation}
%and any such $(m,0)$ is a stationary point of $e$.
%\end{theorem}

@@

\subsection{Proof of Theorem \ref{thm:norminv}}

@@

\subsection{Proof of Theorem \ref{thm:boxcarbad}}
